---
output: pdf_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, autodep=TRUE)
```


Include libraries
```{r}
library(boot)
library(invgamma)
library(mcmc)
library(coda)
library(ggplot2)
```


Task 5
First, we create a help function that divides the number of coal-mining disasters into two, in the intervals $[t_0,t_1)$ and $[t_1,t_2]$.

```{r}
divide_data = function(t) {
  coalEvents = coal[2:190,]
  indicator = coalEvents < t
  index = sum(indicator)
  return (index)
}
```

In order to sample from $t_1|\lambda_0,\lambda_1,\beta,x$, we will use a random walk proposal. We propose a $\tilde{t}|t_1$ from a normal distribution with mean $t_1$, i.e. the last step, and tuning parameter $\sigma$, $\tilde{t}|t_1 \sim N(t_1,\sigma^2)$. The $Q$'s in the acceptance probability term cancels so that the acceptance probability becomes

$$
\alpha = \min\Big(1,\frac{f(\tilde t)}{f(t)} \cdot \frac{Q( t)}{Q(\tilde t)}\Big) =  \min\Big(1,\frac{f(\tilde t)}{f(t)}\Big)
$$
where 
$$
\frac{f(\tilde t | \tilde z)}{f(t| z)} = \frac { \frac{(\lambda_0 (\tilde t_1 - t_0))^{\tilde z_0}}{\tilde{z}_0!} \cdot\frac{(\lambda_1(t_2- \tilde t_1))^{\tilde z_1}}{\tilde{z}_1!} \cdot e^{-\lambda_0(\tilde t_1-t_0)} \cdot e^{-\lambda_1(t_2- \tilde t_1)}}{\frac{(\lambda_0 (t_1 - t_0))^{z_0}}{z_0!} \cdot\frac{(\lambda_1(t_2-t_1))^{z_1}}{z_1!} \cdot e^{-\lambda_0(t_1-t_0)} \cdot e^{-\lambda_1(t_2-t_1))}}
$$
The MCMC algorithm is shown below. Note that the acceptance probability is computed on log-scale.

```{r}
# intital cond
n = 100000

t_0 = coal[1,]
t_2 = coal[191,]
lambda0 = rep(0,n)
lambda1 = rep(0,n)
beta = rep(0,n)
beta[1] = 1
t= rep(0,n)
sigma = (t_2-t_0)/7
#sigma = 8
t[1] = (t_0+t_2)/2
lambda0[1] = 3
lambda1[1] =1
N0 = rep(0,n)
N1 = rep(0,n)
index = divide_data(t[1])
N0[1] = index
N1[1] = 189-index
counterAccRate = 0
fraction_vec = rep(0,1)


#Metropolis Hastings, we know how to sample beta, and lambda, and always accept the proposals. For t, using random walk proposals.

for (i in 1:(n-1)) {
  #i = 1
  t_prop = rnorm(1,t[i],sigma)
  #find split of N_prop
  index = divide_data(t_prop)
  N0_prop = index
  N1_prop = 189-(index)
  
  t_out_of_range = (t_prop > t_2) || (t_prop < t_0)
  #find acceptance probability
  if (!t_out_of_range) {
    logterm1 = N0_prop*(log(lambda0[i]) + log(t_prop-t_0)) + N1_prop*(log(lambda1[i]) + log(t_2-t_prop))
    logterm2 = N0[i]*(log(lambda0[i]) + log(t[i]-t_0)) + N1[i]*(log(lambda1[i]) + log(t_2-t[i]))
    logterm3 = (lambda1[i] -lambda0[i])*(t_prop-t[i])
    logterm4 = lfactorial(N0_prop) + lfactorial(N1_prop)
    logterm5 = lfactorial(N0[i]) + lfactorial(N1[i])
    fraction = (logterm1 - logterm2 + logterm3 - logterm4 + logterm5)
    fraction_vec[i] = fraction
  }
  else {
    fraction = -Inf
  }
  
  alpha = min(0,fraction)
  u = log(runif(1))
  if (u < alpha) { # if true accept new proposal
    N0[i+1] = N0_prop
    N1[i+1] = N1_prop
    t[i+1] = t_prop
    counterAccRate = counterAccRate + 1
  }
  else { #or if false, keep the old state
    N0[i+1] = N0[i]
    N1[i+1] = N1[i]
    t[i+1] = t[i]
  }  
  
  #sample beta
  beta[i+1] = rinvgamma(1,shape = 4, scale = 1/(lambda0[i] + lambda1[i] + 1))
  #sample lambda
  lambda0[i+1] = rgamma(1,shape = N0[i+1] + 2,scale = 1/((t[i+1]-t_0)+1/beta[i+1]))
  lambda1[i+1] = rgamma(1,shape = N1[i+1] + 2,scale = 1/((t_2-t[i+1]) + 1/beta[i+1]))
  
  
}
counterAccRate/n
iter = seq(1,n,1)
plot(iter,t)
ggplot(data = as.data.frame(t), aes(x = iter, y = t) ) + geom_path()
mean(lambda0)
mean(lambda1)
```

Task 8
a)
A block proposal for $( t_1, \lambda_0, \lambda_1$ keeping $\beta$ unchanged. Generate the potential new values
$( \tilde t_1, \tilde \lambda_0, \tilde \lambda_1$ by first generating $\tilde t_1$ from a normal distributi1on centered at the current value of $t_1$ and thereafter generate $\tilde \lambda_0, \tilde \lambda_1$ from their joint full conditionals inserted the potential new
value $t_1$, i.e. $f( \lambda_1, \lambda_2|x, t_1, \beta).$
b)
Now the acceptance probability becomes
where 
$$
\frac{f(\tilde t | \tilde z)}{f(t| z)} = \frac {(\tilde \lambda_0 (\tilde t_1 - t_0))^{\tilde 
z_0} \cdot( \tilde \lambda_1(t_2- \tilde t_1))^{\tilde z_1} \cdot exp( \tilde \lambda_0(\tilde t_1-t_0)) \cdot exp( \tilde \lambda_1(t_2- \tilde t_1))}{(\lambda_0 (t_1 - t_0))^{z_0} \cdot(\lambda_1(t_2-t_1))^{z_1} \cdot exp(\lambda_0(t_1-t_0)) \cdot exp(\lambda_1(t_2-t_1))}
$$
And 
$$
\frac{Q( \lambda_0, \lambda_1,  t_1 |\tilde \lambda_0, \tilde \lambda_1, \tilde t_1) }{Q(\tilde \lambda_0, \tilde \lambda_1, \tilde t_1 |\lambda_0, \lambda_1, t_1)} = \frac{Q(\lambda_0) \cdot Q(\lambda_1) \cdot Q(t_1)}{Q(\tilde \lambda_0) \cdot Q( \tilde \lambda_1) \cdot Q( \tilde t_1)} = \frac{Q(\lambda_0) \cdot Q(\lambda_1)}{Q(\tilde \lambda_0) \cdot Q( \tilde \lambda_1)} = \frac{\lambda_0 exp(\frac{\lambda_0}{\beta}) \cdot \lambda_1 exp(\frac{\lambda_1}{\beta}) }{ \tilde \lambda_0 exp(\frac{ \tilde \lambda_0}{\beta}) \cdot \tilde  \lambda_1 exp(\frac{  \tilde \lambda_1}{\beta}) }
$$
